### 2016-09-20 Projects/DelCo_2014_2015/Reports/Hose_Stream_Tests/TeX_Files/DelCo_Hose_Stream_Tests.tex
Minor changes to hose stream report, only missing TN#

Fire suppression tactics using hose streams can affect ventilation in a structure and may impact the movement of smoke and heat through a structure. Seven experimental series containing a total of 154 different configurations were conducted to study the impact of different hose stream and nozzle movement pattern combinations on air movement within residential scale structures. The experiments studied four different hose streams: a straight stream, narrow fog stream, a wide fog stream from a combination nozzle, and a solid stream from a smooth bore nozzle. The streams were applied from a static, or fixed, position; by moving the hoseline left-to-right across the room in a sweeping motion; and by rotating the hoseline in both the clockwise and counterclockwise directions. Gas velocity was measured at different locations in the structure during the experiments. The wide fog stream caused the most air movement out of any of the tested streams, reaching a maximum velocity of 2.6~m/s (5.8~mph) and maximum air flow rate of approximately 4.2~m$^3$/s (9000~cfm), followed by the narrow fog stream, and then the straight stream and solid stream from the smooth bore nozzle, which both caused approximately the same amount of air movement through the structure. The data were consistent with a straight stream only causing air movement through the structure when it was applied in a moving pattern. Furthermore, there were no statistically significant differences in the average measured air velocity between the clockwise and counterclockwise nozzle movement patterns. It was determined that the type of hose stream and manner in which it is applied dictates the extent to which a stream impacts the ventilation of a structure. To better quantify the impact of such air flows on the fire environment, additional experiments need to be conducted with structure fires in a controlled environment.

@@

There exists a multitude of different tools, methods, and techniques used to apply water for fire suppression using hose streams. A common type of nozzle that is used to apply water during fire suppression in residential structures is an adjustable-pattern spray nozzle, commonly referred to as a combination nozzle~\cite{NFPA_1964}. Most combination nozzles contain an adjustable tip that can be rotated to produce different stream patterns. The types of streams produced by a combination nozzle are often divided into three categories based on the angle of the stream with respect to the centerline of the nozzle: a straight stream typically produces an angle less than 15\SI{}{\degree}; a narrow fog stream produces an angle between 15\SI{}{\degree} and 45\SI{}{\degree}; and a wide fog stream produces an angle between 45\SI{}{\degree} and 80\SI{}{\degree}~\cite{IFSTA:Essentials_of_FF}. Increasing the angle of the stream allows for a larger area to be covered by the water spray. However, increasing the angle decreases the distance reached by the stream and its penetration power.

@@

Four types of hose streams and four nozzle movement patterns were studied under 154 different configurations during the seven test series described in this report. Three of the hose stream patterns studied were from a combination nozzle: straight stream, narrow fog stream, and wide fog stream. Water was flowed into the structure by having the hose in a static, or fixed, position; by moving the hoseline left-to-right across the room in a sweeping motion; and by rotating the hoseline in both the clockwise and counterclockwise directions. Additionally, experiments comparing the differences in impact on air flows in a structure from a combination nozzle straight stream and a solid stream from a smooth bore nozzle equipped with a 1~in. tip were conducted. Images of the four types of hose streams are presented in Fig.~\ref{fig:hose_streams}.

@@

The second floor of the Two Story Structure was built on the wood ceiling support described above and was connected to the first floor by a stairwell. The walls of the second floor were wood-frame with nominal 51~mm (2~in) by 102~mm (4~in) studs set to 400~mm (16~in) centers. The interior walls were protected by 16~mm (0.63~in) fire rated gypsum board, 16~mm (0.63~in) durarock board, and a second layer of 16~mm (0.63~in) fire rated gypsum board. The exterior walls were protected with 11~mm (0.44~in) thick OSB and 8~mm (0.31~in) fiber cement lap siding.

\subsubsection{Layout}
\label{sec:layout}
Simple floor plans showing the total length and width of the Single Story and Two Story Structures are presented in Fig.~\ref{fig:east_dimensioned_plan} and Fig.~\ref{fig:west_dimensioned_plan}, respectively. Fully dimensioned floor plan drawings are provided in Appendix~\ref{chap:floor_plans}.

@@

The exterior doorways of each structure and the stairwell doorway on the second level of the Two Story Structure all contained doors that were opened or closed at certain instances during tests to change the ventilation pathways within the structure. All other doorways in the structures did not contain a door. In order to close these doorways, a sheet of gypsum board was used to cover the opening, and the doorway remained closed for the duration of the test procedure.
\clearpage

\begin{figure}[!ht]
	\includegraphics[width=\columnwidth]{../Figures/Floor_Plans/East_Test_Structure_Dimensioned_LxW}
	\caption[Simple floor plan of the Single Story Structure.]{Simple floor plan of the Single Story Structure.}
	\label{fig:east_dimensioned_plan}
\end{figure}

\begin{figure}[!ht]
	\includegraphics[width=.985\columnwidth]{../Figures/Floor_Plans/West_Test_Structure_2nd_Floor_Dimensioned_LxW}
	\\~\\
	\includegraphics[width=\columnwidth]{../Figures/Floor_Plans/West_Test_Structure_1st_Floor_Dimensioned_LxW}
	\caption[Simple floor plan of the first and second floors of the Two Story Structure.]{Simple floor plan of the second floor (top) and first floor (bottom) of the Two Story Structure.}
	\label{fig:west_dimensioned_plan}
\end{figure}

@@

Three different structure flow path configurations were used for the seven experimental series. Tests~1 and 5 were conducted in the Single Story Structure and used a ``U-shaped'' configuration shown in Fig.~\ref{fig:east_setup}. Test~1 used a monitor and combination nozzle to flow water, and Test~5 used a combination nozzle attached directly to a handline to flow water.

Two different flow path configurations were used for the test series conducted in the Two Story Structure. The main difference between the two flow paths was the position of the south side door on the first level. Tests~2 and 6 used a configuration with the door opened (Fig.~\ref{fig:flow_path_1}), while Tests~3, 4, and 7 used a configuration with the door closed (Fig.~\ref{fig:flow_path_2}). Tests~2 and 3 used a combination nozzle attached to a monitor to flow water; Tests~6 and 7 used a combination nozzle attached directly to a handline to flow water; and Test~4 used both a combination nozzle and a 1~in. tip smooth bore nozzle attached to a monitor to flow water in a straight and solid stream, respectively.

@@

\subsubsection{Test~1}
Test~1 utilized a monitor equipped with a combination nozzle to flow water from Room~C of the Single Story Structure to the area above the doorway on the south wall of Room~B (Fig.~\ref{fig:test_1_pic}). A plan view of the structure's layout during Test~1 can be found in Fig.~\ref{fig:east_setup}. The figure highlights the general direction of the established flow path along with the approximate location of A6, the bi-directional probe array used to measure gas velocity through the northeast interior doorway. The procedure for Test~1 was the simplest of all the test procedures. First, water was flowed in a straight stream for 30 seconds, then the stream was changed to a narrow fog for 30 seconds, and finally the stream was changed to a wide fog for 30 seconds. The procedure was repeated.

\begin{figure}[!ht]
	\includegraphics[trim=16cm 6.25cm 9cm 6cm, clip=true, width=4in]{../Figures/Pictures/SS_Room_B_Test_33}
	\\~\\
	\includegraphics[trim=16cm 6.25cm 9cm 6cm, clip=true, width=4in]{../Figures/Pictures/NF_Room_B_Test_33}
	\\~\\
	\includegraphics[trim=16cm 6.25cm 9cm 6cm, clip=true, width=4in]{../Figures/Pictures/WF_Room_B_Test_33}
	\caption[Straight stream, narrow fog stream, and wide fog stream during Test~1.]{Water flowing from the monitor in Room C in a straight stream (top), narrow fog stream (middle), and wide fog stream (bottom) aimed at the area above the doorway on the south wall of Room B during Test~1 in the Single Story Structure.}
	\label{fig:test_1_pic}
\end{figure}

@@

\begin{figure}[!ht]
	\minipage{2.15in}
	\begin{center}
		\includegraphics[width=2in]{../Figures/Pictures/SS_70}
	\end{center} 
	\endminipage
	\minipage{2.15in}
	\begin{center}
		\includegraphics[width=2in]{../Figures/Pictures/SB_70}
	\end{center}
	\endminipage
	\caption[Straight stream from combination nozzle and solid stream from smooth bore nozzle with 1~in. tip during Test~4.]{Straight stream flowing from the combination nozzle (left) and solid stream flowing from the smooth bore nozzle (right) during Test~4.}
	\label{fig:test_4_pic}
\end{figure}
\FloatBarrier

@@

\subsubsection{Test~5}
Test~5 utilized a 1.75~in. handline equipped with a combination nozzle to flow water from Room C of the Single Story Structure into Room B (Fig.~\ref{fig:test_5_pic}). The layout of the structure was identical to the layout for Test~1, described above in Section~\ref{sec:monitor_procedure} and presented in Fig.~\ref{fig:east_setup}. The procedure for Test~5 began by flowing water in a straight stream aimed at the ceiling of Room B for 30 seconds in a fixed position. After 30 seconds, the nozzle was rotated in the clockwise direction for 30 seconds. Then, the nozzle was aimed at the south doorway of Room B, and water was applied for 30 seconds in a fixed position followed by 30 seconds in a clockwise pattern. The procedure was repeated using a narrow fog stream and a wide fog stream. This entire process was repeated for the streams being applied in the fixed and counterclockwise patterns. The average rotation rate for configurations that involved nozzle rotation during Test~5 was 2.1 rotations per second with a maximum rate of 2.3 rotations per second and minimum of 1.9 rotations per second.

\begin{figure}[!ht]
	\includegraphics[trim=23cm 6.5cm 4cm 6cm, clip=true, width=3.75in]{../Figures/Pictures/SS_Room_B_Test_34}
	\\~\\
	\includegraphics[trim=23cm 6.5cm 4cm 6cm, clip=true, width=3.75in]{../Figures/Pictures/NF_Room_B_Test_34}
	\\~\\
	\includegraphics[trim=23cm 6.5cm 4cm 6cm, clip=true, width=3.75in]{../Figures/Pictures/WF_Room_B_Test_34}
	\caption[Straight stream, narrow fog stream, and wide fog stream during Test~5.]{Straight stream (top), narrow fog stream (middle), and wide fog stream (bottom) aimed at the south doorway of Room B during Test~5.}
	\label{fig:test_5_pic}
\end{figure}

\subsubsection{Tests~6 \& 7}
Tests~6 and 7 followed identical procedures that involved using a 1.75~in. handline equipped with a combination nozzle to flow water from the north side double doors of the Two Story Structure into the first floor room. The water stream was aimed at the south side wall of the first floor during the experiments. Test~6 contained the flow path configuration presented in Fig.~\ref{fig:flow_path_1}, and Test~7 contained the flow path configuration presented in Fig.~\ref{fig:flow_path_2}. Each test began by using a straight stream to flow water in a fixed, horizontal position. After 60 seconds of water flow, the stairwell door was opened. One minute later, the west double door on the north side of the second floor was opened, and water continued to flow for 60 seconds. Next, the water flow was stopped, the two doors were closed, and the procedure was repeated three additional times for the sweeping (back and forth, left-to-right), clockwise, and counterclockwise nozzle motion patterns. This entire process was repeated using narrow fog and wide fog streams. Fig.~\ref{fig:test_6_pic} contains an image of a straight stream pattern being applied during Test~6 with the flow path fully established. During configurations involving nozzle rotation with a wide fog stream or narrow fog stream, the average rotation rate was 1.3 rotations per second with a maximum of 1.4 rotations per second and a minimum of 1.2 rotations per second. The rotation rates for the clockwise and counterclockwise rotation of the straight stream were 0.8 rotations per second and 0.7 rotations per second, respectively, for Test~6, and 0.4 rotations per second for both directions during Test~7.

@@

% ==========================
% = RESULTS AND DISCUSSION =
% ==========================
\chapter{Results and Discussion}
\label{chap:results}
In the following sections, the measurements are presented in graphic and tabular form. In the graphs, an error bar will represent the estimated uncertainty of the measurement. In the tables, the uncertainty will be included in the caption of each table as a percentage enclosed in brackets.

@@

\multirow{5}{*}{\textit{Straight}} & \multirow{2}{*}{\small Velocity} 

@@

\multirow{5}{*}{\begin{tabular}{@{}c@{}} \textit{Narrow} \\ \textit{Fog} \end{tabular}} & \multirow{2}{*}{\small Velocity}

@@

\multirow{5}{*}{\begin{tabular}{@{}c@{}} \textit{Wide} \\ \textit{Fog} \end{tabular}} & \multirow{2}{*}{\small Velocity}

@@

With the exception of the narrow fog stream in the fixed position during Test~6, every hose stream and nozzle movement pattern combination using a narrow fog or wide fog stream during the handline experiments produced air movement---an average gas velocity greater than 0.5~m/s (1.1~mph)---through the interior doorway while the flow path was fully established. For the straight stream, however, air movement through the interior doorway only occurred when the stream was applied in a pattern that involved moving the handline. Similar to the monitor experiments, the average air velocity through the interior doorway was always greater in magnitude for the nozzle movement patterns using the narrow and wide fog streams than the same patterns using the straight stream. Another important result seen in Tables~\ref{table:east_hand_A6_avgs} and \ref{table:west_hand_A10_avgs} is that there are no statistically significant differences (differences greater than the error associated with the bi-directional probes) in the amount of air flow through the interior doorway when rotating the hoseline in the clockwise direction compared to rotating it in the counterclockwise direction for any of three hose streams tested. The similarities in air flow produced by the clockwise and counterclockwise nozzle movement patterns can also be seen graphically in Fig.~\ref{fig:Test_6_BDP_A10_Avg_All} and Figs.~\ref{fig:Test_5_BDP_A6_Avg_All} and \ref{fig:Test_7_BDP_A10_Avg_All} in Appendix~\ref{chap:handline_plots}.

@@

Seven experimental test series containing a total of 154 different configurations were conducted to study the impact of different hose stream and nozzle movement pattern combinations on air movement within residential scale structures.

Three experimental series studied the differences between a straight stream, narrow fog stream, and wide fog stream emitted from a combination nozzle attached to a monitor aimed at specific locations in a structure for different ventilation configurations. For each of the three series, the average air velocity through the interior doorway while the flow path was fully established was greatest for the wide fog stream no matter the location and ventilation pattern within the structure. The magnitude of the average air velocity through the interior doorway while the flow path was fully established for the narrow fog stream was always equal to or greater than the magnitude of the average air velocity for the straight stream.

Additionally, a fourth experimental series using a monitor was conducted in which a straight stream from a combination nozzle was compared to a solid stream from a smooth bore nozzle with a 1~in. tip aimed at a specified location for 60 seconds with a fully established flow path. The streams were compared for six different 60 second periods during which the flow path was fully established: three times each for the monitor aimed at two locations. The results from the four monitor test series suggest a wide fog stream impacts air movement within a structure the most out of any of the tested streams, followed by the narrow fog stream, and then the straight stream and solid stream from a smooth bore nozzle, which both caused approximately the same amount of air movement through the structure.

In addition to the monitor test series, three different experimental series were conducted using a handline to study the impact of different stream and nozzle movement pattern combinations on air movement within a structure. Similar to the monitor experiments, the average air velocity through the interior doorway of a structure with a fully established flow path was greater for the wide fog and narrow fog streams than the straight stream. It was discovered that a straight stream increased air movement through the interior doorway when it was applied in a moving pattern. With regard to Royer's theory, throughout all handline test series, there were no statistically significant differences in the average air velocity measured through the interior doorway between the clockwise and counterclockwise nozzle movement patterns. However, these experiments only addressed cold flow.

For this limited series of water flow experiments, the data demonstrate that for these flows and configurations, hose streams can affect air movement within a structure. The type of hose stream used and manner in which it is applied dictates the extent to which a stream impacts the ventilation of a structure. These experiments have shown there is a potential for hose streams to affect air flows within a structure during fire scenarios. To better quantify the impact of air flows from hose streams on the fire environment, additional experiments need to be conducted with structure fires in a controlled environment.

@@

\appendix
\chapter{Fully Dimensioned Floor Plans}
\label{chap:floor_plans}

@@

\begin{figure}[!ht]
	\includegraphics[width=0.86\columnwidth]{../Figures/Plots/HOSE_IXAOXX_BDP_A6_stream_avgs}
	\caption{Average velocity through stairwell door during Test~5 for three hose stream patterns.}
	\label{fig:Test_5_BDP_A6_Avg_All}
\end{figure}
\FloatBarrier

>>>>

Fire suppression tactics using hose streams can affect ventilation in a structure and may impact the movement of smoke and heat through a structure. Seven experimental series containing a total of 154 configurations were conducted to study the impact of different hose stream and nozzle movement pattern combinations on air movement within residential scale structures. The experiments studied four different hose streams: a straight stream, narrow fog stream, and a wide fog stream from a combination nozzle and a solid stream from a smooth bore nozzle. The streams were applied from a static, or fixed, position; by moving the hoseline left-to-right across the room in a sweeping motion; and by rotating the hoseline in both the clockwise and counterclockwise directions. Gas velocity was measured at different locations in the structure during the experiments. The wide fog stream caused the most air movement out of any of the tested streams, reaching a maximum velocity of 2.6~m/s (5.8~mph) and maximum air flow rate of approximately 4.8~m$^3$/s (10250~cfm), followed by the narrow fog stream, and then the straight stream and solid stream from the smooth bore nozzle, which both caused approximately the same amount of air movement through the structure. The data were consistent with a straight stream only causing air movement through the structure when it was applied in a moving pattern. Furthermore, there were no statistically significant differences in the average measured air velocity between the clockwise and counterclockwise nozzle movement patterns. It was determined that the type of hose stream and manner in which it is applied dictates the extent to which a stream impacts the ventilation of a structure. To better quantify the impact of such air flows on the fire environment, additional experiments need to be conducted with structure fires in a controlled environment.

@@

There exists a multitude of different tools, methods, and techniques used to apply water for fire suppression using hose streams. A common type of nozzle that is used to apply water during fire suppression in residential structures is an adjustable-pattern spray nozzle, commonly referred to as a combination nozzle~\cite{NFPA_1964}. Most combination nozzles contain an adjustable tip that can be rotated to produce different stream patterns. The types of streams produced by a combination nozzle are often divided into three categories based on the angle of the stream with respect to the centerline of the nozzle: a straight stream typically produces an angle less than 15\SI{}{\degree}; a narrow fog stream produces an angle between 15\SI{}{\degree} and 45\SI{}{\degree}; and a wide fog stream produces an angle between 45\SI{}{\degree} and 80\SI{}{\degree}~\cite{IFSTA:Essentials_of_FF}. Increasing the angle of the stream allows for a larger area to be covered by the water spray but also decreases the distance reached by the stream and its penetration power.

@@

Four types of hose streams and four nozzle movement patterns were studied under 154 different configurations during the seven test series described in this report. Three of the hose stream patterns studied used a combination nozzle: straight stream, narrow fog stream, and wide fog stream. Water was flowed into the structure by having the hose in a static, or fixed, position; by moving the hoseline left-to-right across the room in a sweeping motion; and by rotating the hoseline in both the clockwise and counterclockwise directions. Additionally, experiments comparing the differences in impact on air flows in a structure between a straight stream from a combination nozzle and a solid stream from a smooth bore nozzle equipped with a 1~in. tip were conducted. Images of the four types of hose streams are presented in Fig.~\ref{fig:hose_streams}.

@@

The second floor of the Two Story Structure was built on the wood ceiling support described above and was connected to the first floor by a stairwell. The walls of the second floor were wood-frame with nominal 51~mm (2~in) by 102~mm (4~in) studs set to 400~mm (16~in) centers. The interior walls were protected by 16~mm (0.63~in) fire rated gypsum board, 16~mm (0.63~in) durarock board, and a second layer of 16~mm (0.63~in) fire rated gypsum board. The interior ceiling of the second story was covered by two layers of 13~mm (0.5~in) thick cement board. The exterior walls were protected with 11~mm (0.44~in) thick OSB and 8~mm (0.31~in) fiber cement lap siding.

\subsubsection{Layout}
\label{sec:layout}
Simple floor plans with the total length and width of the Single Story and Two Story Structures are presented in Fig.~\ref{fig:simple_east_dimensioned_plan} and Fig.~\ref{fig:simple_west_dimensioned_plan}, respectively. Fully dimensioned floor plan drawings are provided in Appendix~\ref{chap:floor_plans}.

@@

The exterior doorways of each structure and the stairwell doorway on the second level of the Two Story Structure all contained doors that were opened or closed at certain instances during tests to change the ventilation pathways within the structure. All other doorways in the structures did not contain a door. In order to close these doorways, a sheet of gypsum board was used to cover the opening, and the doorway remained closed for the duration of the test procedure.

\begin{figure}[!ht]
	\includegraphics[width=\columnwidth]{../Figures/Floor_Plans/East_Test_Structure_Dimensioned_LxW}
	\caption[Simple floor plan of the Single Story Structure.]{Simple floor plan of the Single Story Structure.}
	\label{fig:simple_east_dimensioned_plan}
\end{figure}

\begin{figure}[!ht]
	\includegraphics[width=.985\columnwidth]{../Figures/Floor_Plans/West_Test_Structure_2nd_Floor_Dimensioned_LxW}
	\\~\\
	\includegraphics[width=\columnwidth]{../Figures/Floor_Plans/West_Test_Structure_1st_Floor_Dimensioned_LxW}
	\caption[Simple floor plan of the first and second floors of the Two Story Structure.]{Simple floor plan of the second floor (top) and first floor (bottom) of the Two Story Structure.}
	\label{fig:simple_west_dimensioned_plan}
\end{figure}

@@

Three different structure flow path configurations were used for the seven experimental series. Tests~1 and 5 were conducted in the Single Story Structure and used the ``U-shaped'' configuration shown in Fig.~\ref{fig:east_setup}. Test~1 used a monitor and combination nozzle to flow water, and Test~5 used a combination nozzle attached directly to a handline to flow water. The two additional flow path configurations were used for the test series conducted in the Two Story Structure. The main difference between the two configurations was the position of the south side door on the ground level. Tests~2 and 6 used a configuration with the door opened (Fig.~\ref{fig:flow_path_1}), while Tests~3, 4, and 7 used a configuration with the door closed (Fig.~\ref{fig:flow_path_2}). Tests~2 and 3 used a combination nozzle attached to a monitor to flow water; Tests~6 and 7 used a combination nozzle attached directly to a handline to flow water; and Test~4 used both a combination nozzle and a 1~in. tip smooth bore nozzle attached to a monitor to flow water in a straight and solid stream, respectively.

@@

\subsubsection{Test~1}
Test~1 utilized a monitor equipped with a combination nozzle to flow water from Room~C of the Single Story Structure to the area above the doorway on the south wall of Room~B (Fig.~\ref{fig:test_1_pic}). A plan view of the structure's layout during Test~1 can be found in Fig.~\ref{fig:east_setup}. The figure highlights the general direction of the established flow path along with the approximate location of A6, the bi-directional probe array used to measure gas velocity through the northeast interior doorway. The procedure for Test~1 was the simplest of all the test procedures. First, water was flowed in a straight stream for 30 seconds, then the stream was changed to a narrow fog for 30 seconds, and finally the stream was changed to a wide fog for 30 seconds. Then, the procedure was repeated an additional time.

\begin{figure}[!ht]
	\includegraphics[trim=16cm 6.25cm 9cm 6cm, clip=true, width=4in]{../Figures/Pictures/SS_Room_B_Test_33}
	\\~\\
	\includegraphics[trim=16cm 6.25cm 9cm 6cm, clip=true, width=4in]{../Figures/Pictures/NF_Room_B_Test_33}
	\\~\\
	\includegraphics[trim=16cm 6.25cm 9cm 6cm, clip=true, width=4in]{../Figures/Pictures/WF_Room_B_Test_33}
	\caption[Straight stream, narrow fog stream, and wide fog stream during Test~1.]{Straight stream (top), narrow fog stream (middle), and wide fog stream (bottom) from the monitor in Room C aimed at the area above the doorway on the south wall of Room B during Test~1 in the Single Story Structure.}
	\label{fig:test_1_pic}
\end{figure}

@@

\begin{figure}[!ht]
	\minipage{2.15in}
	\begin{center}
		\includegraphics[width=2in]{../Figures/Pictures/SS_70}
	\end{center} 
	\endminipage
	\minipage{2.15in}
	\begin{center}
		\includegraphics[width=2in]{../Figures/Pictures/SB_70}
	\end{center}
	\endminipage
	\caption[Straight stream from combination nozzle and solid stream from smooth bore nozzle with 1~in. tip during Test~4.]{Straight stream flowing from the combination nozzle (left) and solid stream flowing from the 1~in. tip of the smooth bore nozzle (right) during Test~4 in the Two Story Structure.}
	\label{fig:test_4_pic}
\end{figure}
\FloatBarrier

@@

\subsubsection{Test~5}
Test~5 utilized a 1.75~in. handline equipped with a combination nozzle to flow water from Room C of the Single Story Structure into Room B (Fig.~\ref{fig:test_5_pic}). The layout of the structure was identical to the layout for Test~1 described above in Section~\ref{sec:monitor_procedure} and presented in Fig.~\ref{fig:east_setup}. The procedure for Test~5 began by flowing water in a straight stream aimed at the ceiling of Room B for 30 seconds in a fixed position. After 30 seconds, the nozzle was rotated in the clockwise direction for 30 seconds. Then, the nozzle was aimed at the south doorway of Room B, and water was applied for 30 seconds in a fixed position followed by 30 seconds in a clockwise pattern. The procedure was repeated using a narrow fog stream and a wide fog stream. This entire process was repeated for the streams being applied in the fixed and counterclockwise patterns. The average rotation rate for the configurations that involved nozzle rotation was 2.1 rotations per second with a maximum rate of 2.3 rotations per second and minimum of 1.9 rotations per second.

\begin{figure}[!ht]
	\includegraphics[trim=23cm 6.5cm 4cm 6cm, clip=true, width=3.70in]{../Figures/Pictures/SS_Room_B_Test_34}
	\\~\\
	\includegraphics[trim=23cm 6.5cm 4cm 6cm, clip=true, width=3.70in]{../Figures/Pictures/NF_Room_B_Test_34}
	\\~\\
	\includegraphics[trim=23cm 6.5cm 4cm 6cm, clip=true, width=3.70in]{../Figures/Pictures/WF_Room_B_Test_34}
	\caption[Straight stream, narrow fog stream, and wide fog stream during Test~5.]{Straight stream (top), narrow fog stream (middle), and wide fog stream (bottom) from the handline in Room C aimed at the south doorway of Room B during Test~5 in the Single Story Structure.}
	\label{fig:test_5_pic}
\end{figure}

\subsubsection{Tests~6 \& 7}
Tests~6 and 7 followed identical procedures that used a 1.75~in. handline equipped with a combination nozzle to flow water from the north side double doors of the Two Story Structure into the first floor compartment. The water stream was aimed at the south side wall of the first floor during the experiments. Test~6 contained the flow path configuration presented in Fig.~\ref{fig:flow_path_1}, and Test~7 contained the flow path configuration presented in Fig.~\ref{fig:flow_path_2}. Each test began by using a straight stream to flow water in a fixed, horizontal position. After 60 seconds of water flow, the stairwell door was opened. One minute later, the west double door on the north side of the second floor was opened, and water continued to flow for 60 seconds. Next, the water flow was stopped, the two doors were closed, and the procedure was repeated three additional times for the sweeping (back and forth, left-to-right), clockwise, and counterclockwise nozzle motion patterns. This entire process was repeated using narrow fog and wide fog streams. Fig.~\ref{fig:test_6_pic} contains an image of a straight stream pattern being applied during Test~6 with the flow path fully established. The average rotation rate for the configurations that involved nozzle rotation with a wide fog stream or narrow fog stream during Tests~6 and 7 was 1.3 rotations per second with a maximum of 1.4 rotations per second and a minimum of 1.2 rotations per second. The average rotation rate for the configurations that involved nozzle rotation with a straight stream was 0.8 rotations per second during Test~6 and 0.4 rotations per second during Test~7.

@@

% ==========================
% = RESULTS AND DISCUSSION =
% ==========================
\chapter{Results and Discussion}
\label{chap:results}
In the following sections, the measurements are presented in graphic and tabular form. In the graphs, an error bar represents the estimated uncertainty of the measurement. In the tables, the uncertainty is included in the caption of each table as a percentage enclosed in brackets.

@@

\multirow{5}{*}{\textit{Straight (SS)}} & \multirow{2}{*}{\small Velocity} 

@@

\multirow{5}{*}{\begin{tabular}{@{}c@{}} \textit{Narrow} \\ \textit{Fog (NF)} \end{tabular}} & \multirow{2}{*}{\small Velocity}

@@

\multirow{5}{*}{\begin{tabular}{@{}c@{}} \textit{Wide} \\ \textit{Fog (WF)} \end{tabular}} & \multirow{2}{*}{\small Velocity}

@@

With the exception of the narrow fog stream in the fixed position during Test~6, every hose stream and nozzle movement pattern combination using a narrow fog or wide fog stream during the handline experiments produced air movement---an average gas velocity greater than 0.5~m/s (1.1~mph)---through the interior doorway while the flow path was fully established. For the straight stream, however, air movement through the interior doorway only occurred when the stream was applied in a pattern that involved moving the nozzle. Similar to the monitor experiments, the average air velocity through the interior doorway was always greater in magnitude for the nozzle movement patterns using the narrow and wide fog streams than the same patterns using the straight stream. Another important result seen in Tables~\ref{table:east_hand_A6_avgs} and \ref{table:west_hand_A10_avgs} is that there are no statistically significant differences (greater than the uncertainty associated with the bi-directional probes) in the amount of air flow through the interior doorway when rotating the hoseline in the clockwise direction compared to rotating it in the counterclockwise direction for any of three hose streams tested. The similarities in air flow produced by the clockwise and counterclockwise nozzle movement patterns can also be seen graphically in Fig.~\ref{fig:Test_6_BDP_A10_Avg_All} and Figs.~\ref{fig:Test_5_BDP_A6_Avg_All} and \ref{fig:Test_7_BDP_A10_Avg_All} in Appendix~\ref{chap:handline_plots}.

@@

Seven experimental test series containing a total of 154 configurations were conducted to study the impact of different hose stream and nozzle movement pattern combinations on air flows within residential scale structures.

Three of the experimental series studied the differences between a straight stream, narrow fog stream, and wide fog stream emitted from a combination nozzle attached to a monitor aimed at specific locations in a structure containing different ventilation configurations. For each of the three test series, the average air velocity through the interior doorway while the flow path was fully established was greatest for the wide fog stream no matter the location and ventilation pattern within the structure. The magnitude of the average air velocity through the interior doorway while the flow path was fully established for the narrow fog stream was always equal to or greater than the magnitude of the average air velocity for the straight stream. Additionally, a fourth experimental series using a monitor was conducted to compare the impact of a straight stream from a combination nozzle to a solid stream from a smooth bore nozzle with a 1~in. tip on air flows within a structure. The streams were compared across six different 60 second periods with a fully established flow path: three times each for the monitor aimed at two locations. The results from the four monitor test series suggest a wide fog stream impacts air movement within a structure the most out of any of the tested streams, followed by the narrow fog stream, and then the straight stream and solid stream from a smooth bore nozzle, which both caused approximately the same amount of air movement through the structure.

In addition to the monitor test series, three different experimental series were conducted using a combination nozzle attached to a handline to study the impact of different stream and nozzle movement pattern combinations on air movement within a structure. Similar to the monitor experiments, the average air velocity through the interior doorway of a structure with a fully established flow path was greater for the wide fog and narrow fog streams than the straight stream. It was discovered that a straight stream increased air movement through the interior doorway only when it was applied in a moving pattern. With regard to Royer's theory, throughout all the handline test series, there were no statistically significant differences in the average air velocity measured through the interior doorway between the clockwise and counterclockwise nozzle movement patterns. However, these experiments only addressed cold flow.

The results from the limited series of water flow experiments demonstrate that hose streams can affect air movement within a structure for the described flows and configurations. The type of hose stream used and manner in which it is applied dictates the extent to which a stream impacts the ventilation of a structure. These experiments have shown there is a potential for hose streams to affect air flows within a structure during fire scenarios. To better quantify the impact of air flows from hose streams on the fire environment, additional experiments need to be conducted with structure fires in a controlled environment.

@@

\appendix
\chapter{Dimensioned Floor Plans}
\label{chap:floor_plans}

@@

\begin{figure}[!ht]
	\includegraphics[width=0.86\columnwidth]{../Figures/Plots/HOSE_IXAOXX_BDP_A6_stream_avgs}
	\caption{Average velocity through stairwell door during Test~5 for the three hose stream patterns.}
	\label{fig:Test_5_BDP_A6_Avg_All}
\end{figure}
\FloatBarrier
